Negative and Fractional Types

### 2012-06-29 rational_ext.tex
negatives

What is a negative type anyway?

@@

For convenience, we sometimes use the notations {{b1 - b2}} and
{{b1/b2}} to indicate the types {{b1 + (-b2)}} and {{b1 * (1/b2)}}
respectively.  The types of the new constructs are:
\vspace{-15pt}
\begin{multicols}{2}  
%subcode{opsem} include main
% eta+ &: 0 <-> (-b) + b :& eps+

@@

%subcode{opsem} include main
%! columnStyle = rclr
%! fwd = \triangleright
%! bck = \triangleleft
% {[eta+, right v, C, s]}^^^ &|-->& {[eta+, left (-v), C ]}
% {[eta+, left (-v), C, s]}^^^ &|-->& {[eta+, right v, C ]}

  \end{enumerate}

@@

\begin{proposition}[Coherence]

>>>>

Charles Pinter in his book on Abstract Algebra talks about how
negatives were used by \emph{logisticians} of a 1000 years ago used
negative numbers as a ``useful fiction''.  The concept of negative
numbers were ``discovered''much later than fractionals. Like Joan Baez
says ``half an apple is easier to undertand than a negative apple''.

Negative types should correspond to sets with negative arity. But what
does that mean? Various approaches have been proposed -- Loeb uses the
notion of multiplicity functions to express negative sized sets
[CITE]. These notions have been extended to Euler characteristic of
categories by Leinster, Joyal and Schanuel -- negative values are
explained sometimes as eqaulities relating otherwise indepedant
objects -- Baez tells a story about ``island'' and positive and
``bridges connecting islands'' as negatives (since the effectively
reduce trhe number of disconnected elements.)

We use the use the syntax {{-b}} for negative types and {{-v}} for
negative values and the axiom {{0 <-> (-b)+b}} as the basis for
negatives. In {{Pi}}/{{PiEE*}}, values flowed from left-to-right in
circuits. Negative types are negative in the sense that they flow in
the oppostite direction.

@@

For convenience, we sometimes use the notations {{b1 - b2}} to
indicate the types {{b1 + (-b2)}}.  The types of the new constructs
are:
\vspace{-15pt}
\begin{multicols}{2}  
%subcode{opsem} include main
% eta+ &: 0 <-> (-b) + b :& eps+

@@

%subcode{opsem} include main
%! columnStyle = rclr
%! fwd = \triangleright
%! bck = \triangleleft
% {[eta+, right v, C, s]}^^^ &|-->& {[eta+, left (-v), C ]}
% {[eta+, left (-v), C, s]}^^^ &|-->& {[eta+, right v, C ]}
  \end{enumerate}

Consider the combinator {{c:(-1+1)+1<->1}} defined below. On providing
the value {{right ():(-1+1)+1}} we get the output {{():1}}. But on
supply input {{left (right ())}} evaluation stop on the reverse
interpreter with value {{left (left ()):(-1+1)+1}}.

By definition, since negative values flow in the opposite direction of
positive values, this is not surprising. Combinators can terminate in
the forward interpreter (with result of type {{b2}}) or in the reverse
interpreter (with result of type {{b1}}). To define an {{eval}} in a
consistent manner without worrying about state of the evaluator, we
transform every combinator of the form {{c:b1<->b2}} to one of the
form {{c:-b2+b1<->0}} by the construction {{action(c)=id(+)c(;)eps+}}.

\begin{definition}[{{eval(c, v)}}]
For {{c:b1<->b2}} and {{v:b1}} we define {{eval(c, v)=v'}} such that
\begin{enumerate}
\item {{v':-b2+b1}} and
\item {{<action(c), right v, []> |-->* <action(c),v',[]>^^^}}
\end{enumerate}
\end{definition}

\begin{definition}[{{c1=c2}}]
  For {{c1:b1<->b2}} and {{c2:b1<->b2}}, we define {{c1=c2}} if
  {{~forall v:b1.eval(c1,v) = eval(c2,v)}}.
\end{definition}

@@

%%%%%%%%%
\subsection{Categorical Structure}
% \label{sec:neg-constructions}

As before, we can show that {{PiEE+}} is a bimonoidal
category. Additionally it has a compact closed structure in the
{{(0,+)}} monoid.

\begin{proposition}[Compact Closed {{(0,+,-)}}]
  {{PiEE+}} is a compact closed category with over the symmetric
  monoid {{(0,+)}} with the dual of objects {{b}} give by {{-b}} with
  morphisms:

\begin{enumerate}
\item {{0 <-> -b + b}} given by {{eta+}}, called the `unit map'.
\item {{b + -b <-> 0}} given by {{swap+ (;) eps+}}, called the `counit map'.
\end{enumerate}

\begin{enumerate}
\item {{zeroi (;) swap+ (;) (id (+) eta+) (;) assocl+ (;) ((swap+ (;) eps+) (+) id) (;) zeroe = id : b <-> b}}
\item {{zeroi (;) (eta+ (+) id) (;) assocr+ (;) (id (+) (swap+ (;) eps+)) (;) swap+ (;) zeroe = id : -b <-> -b}}
\end{enumerate}
